---
title: "HW7"
author: "Joel Essenburg"
format: html
embed-resources: true
code-tools: true
---

```{r}
library(vegabrite)
```

## Exercise 1

### a

- bar charts start at 0

- you can save horizontal space by placing bars closer together and rotating labels, or swapping the x and y axes

- we must pay attention to the order in which the bars are arranged

- we should only rearrange bars when there is no natural ordering to the categories the bars represent  

- you can use a grouped bar chart to visualize two categorical variables at the same time

- showing separate bar plots can be easier to read than grouped bars

- stacking bars is useful when the sum of the amounts represented by the individual stacked bars is a meaningful amount

### b

It is not advised to use a bar chart when the data doesn't need to start at 0 because bar charts need to start at 0 and this can become impractical and obscure key features.

### c

A couple alternatives to bar charts are dot plots and heatmaps. 

### d

Wilke says that stacking bars is more useful when the sum of the amounts represented by the individual stacked bars is in itself a meaningful amount. Stacking bars is also useful when the individual bars represent counts. 

### e

```{r}
as_vegaspec(list(
  `$schema` = "https://vega.github.io/schema/vega-lite/v5.json",
  data = list(
    url = "https://calvin-data304.netlify.app/data/cow-movies.csv" 
  ),
  width = 600,
  height = 400,
  mark = "bar",
  encoding = list(
    x = list(
      field = "amount",
      type = "quantitative",
      title = "Weekend Gross (USD)"
    ),
    y = list(
      field = "title",
      type = "nominal",
      sort = list("-x"),
      title = "Movie"
    )
  )
))
```

### f

```{r}
as_vegaspec(list(
  `$schema` = "https://vega.github.io/schema/vega-lite/v5.json",
  data = list(
    url = "https://calvin-data304.netlify.app/data/cow-income.csv" 
  ),
  width = 800,
  height = 400,
  mark = "bar",
  encoding = list(
    x = list(
      field = "race",
      type = "nominal",
      title = "Race",
      axis = list(labelAngle = 0)
    ),
    y = list(
      field = "median_income",
      type = "quantitative",
      title = "Median Income (USD)"
    ),
    xOffset = list(
      field = "age",
      type = "nominal",
      title = "Age (years)"
    ),
    color = list(
      field = "age",
      type = "nominal",
      title = "Age (years)"
    )
  )
))
```

```{r}
as_vegaspec(list(
  `$schema` = "https://vega.github.io/schema/vega-lite/v5.json",
 data = list(
    url = "https://calvin-data304.netlify.app/data/cow-income.csv" 
  ),
  columns = 3,
  facet = list(
    field = "race",
    type = "nominal",
    title = "Race"
  ),
  spec = list(
    width = 150,
    height = 200,
    mark = "bar",
    encoding = list(
      x = list(
        field = "age",
        type = "ordinal",
        title = "Age Group"
      ),
      y = list(
        field = "median_income",
        type = "quantitative",
        title = "Median Income (USD)"
      ),
      color = list(
        field = "age",
        type = "ordinal",
        title = "Age (Years)",
        legend = NULL
      )
    )
  )
))
```

### g

```{r}
as_vegaspec(list(
  `$schema` = "https://vega.github.io/schema/vega-lite/v5.json",
  data = list(
    url = "https://calvin-data304.netlify.app/data/cow-gapminder.csv" 
  ),
  transform = list(
    list(
      filter = "datum.year == 2007") 
  ),
  width = 600,
  height = 2000,
  mark = "point",
  encoding = list(
    x = list(
      field = "lifeExp",
      type = "quantitative",
      title = "Life Expectancy (Years)"
    ),
    y = list(
      field = "country",
      type = "nominal",
      title = "Country"
    )
  )
))
```

## Exercise 2

### a

Video watched

### b

- the audience should be able to decode the contents of your graph to get an accurate sense of what the original underlying data is and what it is saying

- sometimes it's worth showing the baseline (0), but not always

- a good order to follow is layering information, highlighting the things you want to pay attention to, and repeating design elements

### c

The figures in the video that come from Chapter 1 are the one of Napoleon's army in Russia, and I'm not 100% on this one but I think the second one is the globe map.

## Exercise 3

### a

In Vega-Lite, to make a heat map you must set "mark": "rect".

### b

Figure 6.14

```{r}
as_vegaspec(list(
  `$schema` = "https://vega.github.io/schema/vega-lite/v5.json",
  data = list(
    url = "https://calvin-data304.netlify.app/data/cow-internet2.csv"
  ),
  width = 600,
  height = 400,
  mark = "rect",
  encoding = list(
    x = list(
      field = "year",
      type = "ordinal",
      title = "Year"
    ),
    y = list(
      field = "country",
      type = "nominal",
      sort = list(
        field = "users",
        order = "descending",
        op = "max",
        condition = list(
          field = "year",
          test = "datum.year == 2016" 
        )
      ),
      title = "Country"
    ),
    color = list(
      field = "users",
      type = "quantitative",
      title = "Internet Users / 100 People",
      scale = list(
        scheme = "blues"
      ) 
    )
  )
))
```

### c

A possible problem with heatmaps is that many times it's hard to tell slight differences and similarities in color. 

### d

The problem from earlier is not an issue anymore because the slight differences don't matter as much as the overall and bigger differences. It's a different message that this heatmap is trying to convey, showing more of a bigger picture scenario. It's less of an issue in the figure I made because that figure is also trying to show bigger picture trends as opposed to identifying individual data points as colors.

## Exercise 4

### a

Healy says that pie charts are harder to read because we are better at decoding lengths than decoding angles and relative areas. 

### b

While Healy advocates against pie charts, Wilke is more indifferent, insisting that none of the visualizations are consistently superior over the other, and that there can be situations where a pie chart is a better visualization for a given dataset. For example, Wilke says that pie charts work well when the goal is to emphasize simple fractions, and when we have very small datasets. 

### c

One alternative to pie charts is side-by-side bars. Wilke favors side-by-side bars when highlighting small changes and differences in data and when visualizing the whole isn't as important. Another alternative is stacked bars, which Wilke favors when there are only two bars in the stack.

### d

```{r}
as_vegaspec(list(
  `$schema` = "https://vega.github.io/schema/vega-lite/v5.json",
  data = list(
    values = list(
      list(
        name = "SPD", 
        value = 214
      ),
      list(
        name = "FDP", 
        value = 39
      ),
      list(
        name = "CDU/CSU", 
        value = 243
      )
    )
  ),
  mark = "arc",
  encoding = list(
    theta = list(
      field = "value", 
      type = "quantitative"
    ),
    color = list(
      field = "name", 
      type = "nominal"
    )
  )
))
```




